15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 3

Another category of nuisance species includes mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (Averill *et al* 2018). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

>>>>

Mammals whose population densities have been inflated due to human activities are also considered nuisance species [@murphy_invasive_2014; @clout_invasion_2008], and in United States (US) eastern deciduous forests, this includes white-tailed deer (*Odocoileus virginianus*). Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species [*e.g.*, oak, *Quercus* spp.; @bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
